IML Term project paper

### 2022-12-23 paper.rmd
Update paper.rmd

The key objective in this project, classifying NPF event and nonevent days correctly, is an example of *supervised learning*: our data set contains the correct classes, and we can use these in the training of our models. With *unsupervised learning* we can look for insights into the data without using the known classes. There are many different types of unsupervised learning, discussed in [@book, p.497-552], but since our goal is classification, we focus on *cluster analysis*. This means that we look for subgroups of similar observations. In particular, we use K-means clustering to see if we can easily differentiate between event and nonevent days, and between event types Ia, Ib, II and nonevent, without using the given classes except to validate the results. The K-means clustering algorithm starts by assigning each observation to a cluster at random. At every iteration, the algorithm computes a *cluster centroid*, a vector of the variable means for each observation in the *k*th cluster, and then reassigns each observation to the cluster with the nearest centroid. The nearest centroid is the one to which an observation has the smallest Euclidean distance.

Recall that we have daily means and standard deviations in the dataset. For the purposes of K-means clustering, we restrict ourselves to the means, and because the variables are measured on very different scales, we normalize them to have zero mean and unit variance with the *scale* function. The random cluster assignments tend to have an effect on the end result, so we run the *kmeans* function with 1000 random starting points, and look at the best result. With 2 clusters, the confusion matrix given by the *solve_LSAP* function from the [@clue] package that matches the clusters with the known classes, is as follows:

@@

```{r}

@@

In particular we notice that with several of the variables the boxplots for the nonevent group observations differ more from the observations for days with NPF events than the boxplots for the different NPF event groups both in terms of smaller or higher values but also spanning larger intervals (although this can partially be caused by there being observations within this group). Another notable thing is that the measurements of the same variables at different heights behave very similarly, which imply correlation between them (we take a more detailed look into this in the section *Correlation between parameters*). While looking at similar visualizations of other variables, we noticed a few interesting sets of variables such as nitrogen monoxide (measurements NO), potential temperature gradient (measurements PTG), rain indicator signal (measurements SWS), sulphur dioxide concentration (measurements SO2) and (measurements CS), which seem to mostly have very small standard deviations. This would imply that the shifts during the day in relation to the corresponding mean observations are on average very small.

@@

The correlation between the predictors and the value to be predicted is a very natural starting point when it comes to correlation analysis. It is also possible that the predictors themselves are correlated. Many classifiers are affected by high correlation, i.e. *collinearity* between variables. In regression models, for example, it can be difficult to differentiate between the individual effects of collinear variables on the response. Let's visualize these correlations with R package [@corrplot2021]

@@

Hence we discard most of the measurements of H2O and other variables for which we have multiple highly correlating measurements, both the means and the standard deviations. We keep the parameters measured at 16.8 meters as for some parameters, the measurements have only been measured at that height. This rather crude method of discarding data leaves us with 36 unique columns. The figure below describes the correlation coefficients between the remaining variables

@@

## Description of considered machine learning approaches
In order to use a model that describes the process that creates the data the best, we attempted to use several different kinds of models to assess the data. We test these models on the original dataset or on a suitably modified version of it. After defining each of these models, we would assess which of the models we used performed best and use said model in our predictions.

@@

However, it is not sufficient to simply define the model on the training dataset and assess accuracy and perplexity on said training dataset as this would woefully inflate the accuracy assessment and deflate the perplexity assessment of the model on a general dataset. We instead attempted to assess accuracy and perplexity through cross-validation methods, specifically leave-one-out cross-validation (loo-cv). The idea behind loo-cv according to the course material [@book, p. 200-202] is that for each observation in the training dataset, a model is trained based on all other observations in the dataset. This new model is then used to predict the training dataset value that was left out of defining the model. With this method, we receive based on observations of explanatory variables both predicted values and estimates of probabilities for specific events for each measurement in a way that didn't include the observations in both defining the model training and assessing its quality. Thus, we can use these measures in our accuracy and perplexity calculations to better assess how the model using all the observations in the training dataset behaves on a completely separate test dataset.
```{r preprocess data, echo=FALSE}
# Sample half of the data for training and testing sets
set.seed(42)
idx <- sample(nrow(npfData), nrow(npfData)/2)
npf_train <- npfData[idx, ]
npf_test <- npfData[-idx, ]
```
Next, let's discuss some of the models we considered

@@

After this, we had a remaining set of 18 explanatory variables, which we modelled using logistic regression. After checking the coefficients, there were still variables for which the Wald tests gave relatively high p-values for hypotheses involving non-zero coefficients. In order to deal with this, we set those coefficients equal to zero and assessed the model with the remaining explanatory variables. Through this, we received a simpler model with  explanatory variables.

@@

Lasso regression has obvious links to logistic regression but it includes a penalty term which punishes models with high values of coefficients. According to the course material [@book, p. 241-242] in comparable linear regression minimizing squared error, the minimizable function is the sum of the squared error terms and $\sum_{i=1}^n \gamma |\lambda_i|$ where $\gamma>0$. We similarly as in logistic regression assess coefficients for a model that describes parameter $q$ of a binomial model that minimizes this function.

@@

In logistic regression models, we directly model $p(Y|X)$, i.e. the conditional distribution of the response $Y$ given the predictors $X$. Generative models are covered in [@book, p.141-158]. Generative models offer a less direct approach, where the focus is in modeling $p(X|Y)$ and using these estimates to estimate $p(Y|X)$ for each possible class $k$ with Bayes' theorem:
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X|Y=k)}{\sum_i \pi_i p(X|Y=i)}.
\end{equation*}
Here $pi_k$ is an estimate of the prior probability that a random observation comes from the $k$th class, in our case computed as the fraction of the observations in the training dataset that are in the $k$th class.

@@

#### Linear Discriminant Analysis

@@

In our modeling, we used the function lda from the package [@MASS]. We tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Quadratic Disriminant Analysis

@@

In our approach, we used the qda function from the [@MASS] package, and as with linear discriminant analysis, tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Naive Bayes

A naive Bayes classifier makes no assumptions on the distribution of the observations, and instead assumes that within the *k*th class, the *p* predictors are independent. In other words, the assumption is that there is no association between the predictors, and therefore no joint distribution to consider. With this, the posterior probability can be computed as
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X_1|Y=k) \cdots p(X_p|Y=k)}{\sum_i \pi_i p(X_1|Y=i) \cdots p(X_p|Y=i)}.
\end{equation*}
In a naive Bayes model, there are a few possible ways to estimate $p(X|Y)$. For quantitative predictors such as ours, we assume that within each class, the *j*th predictor is drawn from a univariate normal distribution, i.e. $X_j|Y = k \approx N (\mu_{jk}, \sigma^2_jk)$.
The assumption of independence between predictors is often not realistic, but the model can perform well nonetheless. This is especially true for small datasets, since estimating a joint distribution requires a large amount of data. 
We used the naiveBayes function from the package [@e1071] and the original dataset.

@@

In tree-based methods, the data points are divided into smaller and smaller fragments based on some set of rules. A new division happens at each branch, so when the size of the tree grows, it becomes more and more fitted to the data. When fitting classification trees, the process is often first started with a very large tree, and then the unnecessary branches are removed in the process of pruning, until the complexity of the tree is reduces to the optimal level. Constructing the trees, pruning, and evaluating the results can be done in a variety of different ways.

A often used tree-based classification method is called Random Forest. It is based on the process described above, but only a random subset of the features is considered at each split of the tree. This is repeated multiple times and in the end, the obtained forest is evaluated and the best splits at each point are found. For Random Forest classifier, we used the R library *randomForest*.

@@

The results for the two-class models in terms of loo-cross-validated accuracy and perplexity are:
\begin{table}
\begin{tabular}{|r|ccc|}
\hline
Model type & Loo-CV accuracy & Loo-CV perplexity&Binary accuracy on test data & Binary accuracy on npf_test\\
\hline
Dummy model & 0 & 2.01&\\
Logistic regression & 0.8728448 & 1.338286&0.8445596\\
Naive Bayes & 0.81 & 152.85\\
LDA & 0.89 & 1.33\\
QDA & 0.85 & 6.34\\
SVM & 0.8793103 & -&0.8435233\\
Lasso & 0.8965517 & 1.29399&0.8746114\\
Random Forest & 0.89 & 1.35\\
\hline
\end{tabular}
\end{table}

We also calculated for curiosity's sake the binary accuracies on the npf_test data on some of our models as can be seen above (eventually we came up with the idea very late and thus didn't have time to run it for LDA, QDA and Naive Bayes), even though these results were not part of the process of choosing our model.

@@

With our observed best-performing models being the the random forest model, the Lasso regression model and the linear discriminant analysis. Out of these models, we eventually chose to use the Lasso model as it was the best-performing model. The amount of information we could gain through CV calculations in R without R crashing were relatively limited, to the degree that we couldn't usually even calculate the cross-validated best choice for the tuning parameter. As a result, we chose to forego cross-validation in relation to our multinomial model.

We eventually also chose to not use the LDA model after assessing the assumptions related to the model. LDA models have been found to perform best with a limited amount of variables and we still question whether we should attempt to reduce the amount of variables from the model further from the current amount of 38 from the model. On top of this, considering the assumption of equal variances, the boxplots we drew earlier in the report imply that class "nonevent" might have larger variance for some variables than class "event".

We made the decision of model based on preliminary results, and though classification trees would represent an equally viable solution, we calculated those results only later on, and by that point had already calculated the results for the logistic regression model.

Thus we were left with the Lasso model. However, we could question whether for example specific variables need to be observed based on other research, and Lasso doesn't offer possibilities for this. Because cross-validation with a smaller amount of folds can give varying results different times it has been run, we eventually chose to run the cross-validation several times, check the interval of values received this way, and make decisions based on this interval.

@@

\begin{table}

\begin{tabular}{|r|c|}

\hline

Explanatory variable & Estimate of coefficient\\

\hline

Intercept&16.16886\\

CO2168.std&0.02065910\\

CO242.mean&-0.04024650\\

CO2504.mean&-0.02066992\\

H2O42.mean&-0.2627714\\

H2O84.std&-2.407117e-01\\

NO168.std&2.281875\\

NOx672.std&-0.06064781\\

O3672.mean&0.09233133\\

O384.std&0.03595632\\

Pamb0.std&0.2960355\\

PTG.mean&-15.19914\\

RGlob.mean&0.001578964\\

RHIRGA168.std&0.01765280\\

RHIRGA336.mean&-0.02661884\\

RHIRGA672.mean&-0.03528919\\

RHIRGA672.std&0.06932797\\

SO2168.std&0.7092859\\

SWS.mean&0.01112000\\

T168.std&0.4947142\\

CS.mean&-938.8368\\

CS.std&580.6009\\

\hline

\end{tabular}

\end{table}

@@

## Delivarables

**For the deliverables, we give a grade x.**

>>>>

The key objective in this project, classifying NPF event and nonevent days correctly, is an example of *supervised learning*: our data set contains the correct classes, and we can use these in the training of our models. With *unsupervised learning* we can look for insights into the data without using the known classes. There are many different types of unsupervised learning, discussed in the course material [@book, p.497-552], but since our goal is classification, we focus on *cluster analysis*. This means that we look for subgroups of similar observations. In particular, we use K-means clustering to see if we can easily differentiate between event and nonevent days, and between event types Ia, Ib, II and nonevent, without using the given classes except to validate the results. The K-means clustering algorithm starts by assigning each observation to a cluster at random. At every iteration, the algorithm computes a *cluster centroid*, a vector of the variable means for each observation in the *k*th cluster, and then reassigns each observation to the cluster with the nearest centroid. The nearest centroid is the one to which an observation has the smallest Euclidean distance.

Recall that we have daily means and standard deviations in the dataset. For the purposes of K-means clustering, we restrict ourselves to the means, and because the variables are measured on very different scales, we normalize them to have zero mean and unit variance with the *scale* function. The random cluster assignments tend to have an effect on the end result, so we run the *kmeans* function with 1000 random starting points, and look at the best result. With 2 clusters, the confusion matrix given by the *solve_LSAP* function from the clue[@clue] package that matches the clusters with the known classes, is as follows:

@@

```{r}0.8455959

@@

In particular we notice that with several of the variables the boxplots for the nonevent group observations differ more from the observations for days with NPF events than the boxplots for the different NPF event groups both in terms of smaller or higher values but also spanning larger intervals (although this can partially be caused by there being observations within this group). Another notable thing is that the measurements of the same variables at different heights behave very similarly, which imply correlation between them (we take a more detailed look into this in the section *Correlation between variables*). While looking at similar visualizations of other variables, we noticed a few interesting sets of variables such as nitrogen monoxide (measurements NO), potential temperature gradient (measurements PTG), rain indicator signal (measurements SWS), sulphur dioxide concentration (measurements SO2) and (measurements CS), which seem to mostly have very small standard deviations. This would imply that the shifts during the day in relation to the corresponding mean observations are on average very small.

@@

The correlation between the predictors and the value to be predicted is a very natural starting point when it comes to correlation analysis. It is also possible that the predictors themselves are correlated. Many classifiers are affected by high correlation, i.e. *collinearity* between variables. In regression models, for example, it can be difficult to differentiate between the individual effects of collinear variables on the response. Let's visualize these correlations with R package corrplot[@corrplot2021]

@@

Hence we discard most of the measurements of H2O and other variables for which we have multiple highly correlating measurements, both the means and the standard deviations. We keep the parameters measured at 16.8 meters as for some parameters, the measurements have only been measured at that height. This rather crude method of discarding data leaves us with 38 unique columns. The figure below describes the correlation coefficients between the remaining variables

@@

## Description of considered machine learning approaches

In order to use a model that describes the process that creates the data the best, we attempted to use several different kinds of models to assess the data. We test these models on the original dataset or on a suitably modified version of it. After defining each of these models, we would assess which of the models we used performed best and use said model in our predictions.

@@

However, it is not sufficient to simply define the model on the training dataset and assess accuracy and perplexity on said training dataset as this would woefully inflate the accuracy assessment and deflate the perplexity assessment of the model on a general dataset. We instead attempted to assess accuracy and perplexity through cross-validation methods, specifically leave-one-out cross-validation (loo-cv). The idea behind loo-cv according to the course material [@book, p. 200-202], is that for each observation in the training dataset, a model is trained based on all other observations in the dataset. This new model is then used to predict the training dataset value that was left out of defining the model. With this method, we receive based on observations of explanatory variables both predicted values and estimates of probabilities for specific events for each measurement in a way that didn't include the observations in both defining the model training and assessing its quality. Thus, we can use these measures in our accuracy and perplexity calculations to better assess how the model using all the observations in the training dataset behaves on a completely separate test dataset.
```{r preprocess data, echo=FALSE}
# Sample half of the data for training and testing sets
set.seed(42)
idx <- sample(nrow(npfData), nrow(npfData)/2)
npf_train <- npfData[idx, ]
npf_test <- npfData[-idx, ]
```
Next, let's discuss some of the models we considered

@@

After this, we had a remaining set of 18 explanatory variables, which we modeled using logistic regression. After checking the coefficients, there were still variables for which the Wald tests gave relatively high p-values for hypotheses involving non-zero coefficients. In order to deal with this, we set those coefficients equal to zero and assessed the model with the remaining explanatory variables. Through this, we received a simpler model with  explanatory variables.

@@

Lasso regression has obvious links to logistic regression but it includes a penalty term which punishes models with high values of coefficients. According to the course material [@book, p. 241-242], in comparable linear regression minimizing squared error, the minimizable function is the sum of the squared error terms and $\sum_{i=1}^n \gamma |\lambda_i|$ where $\gamma>0$. We similarly as in logistic regression assess coefficients for a model that describes parameter $q$ of a binomial model that minimizes this function.

@@

In logistic regression models, we directly model $p(Y|X)$, i.e. the conditional distribution of the response $Y$ given the predictors $X$. Generative models are covered in the course material [@book, p.141-158]. Generative models offer a less direct approach, where the focus is in modeling $p(X|Y)$ and using these estimates to estimate $p(Y|X)$ for each possible class $k$ with Bayes' theorem:
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X|Y=k)}{\sum_i \pi_i p(X|Y=i)}.
\end{equation*}
Here $pi_k$ is an estimate of the prior probability that a random observation comes from the $k$th class, in our case computed as the fraction of the observations in the training dataset that are in the $k$th class.

@@

#### Linear Discriminant Analysis \

@@

In our modeling, we used the function lda from the package MASS[@MASS]. We tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Quadratic Disriminant Analysis \

@@

In our approach, we used the qda function from the MASS[@MASS] package, and as with linear discriminant analysis, tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Naive Bayes \

A naive Bayes classifier makes no assumptions on the distribution of the observations, and instead assumes that within the *k*th class, the *p* predictors are independent. In other words, the assumption is that there is no association between the predictors, and therefore no joint distribution to consider. With this, the posterior probability can be computed as
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X_1|Y=k) \cdots p(X_p|Y=k)}{\sum_i \pi_i p(X_1|Y=i) \cdots p(X_p|Y=i)}.
\end{equation*}
In a naive Bayes model, there are a few possible ways to estimate $p(X|Y)$. For quantitative predictors such as ours, we assume that within each class, the *j*th predictor is drawn from a univariate normal distribution, i.e. $X_j|Y = k \approx N (\mu_{jk}, \sigma^2_jk)$.
The assumption of independence between predictors is often not realistic, but the model can perform well nonetheless. This is especially true for small datasets, since estimating a joint distribution requires a large amount of data. 
We used the naiveBayes function from the package e1071[@e1071] and the original dataset.

@@

In tree-based methods, the data points are divided into smaller and smaller fragments based on some set of rules. A new split happens at each node, so when the size of the tree grows, it becomes more and more fitted to the data. When fitting classification trees, the process is often first started with a very large tree, and then the unnecessary branches are removed in the process of pruning, until the complexity of the tree is reduces to the optimal level. Constructing the trees, pruning, and evaluating the results can be done in a variety of different ways.[@book, p. 328-361]

We tried some tree-based models, but didn't have time to finish any of them before submitting the preliminary report. Therefore, we didn't end up selecting any of them as our classifier, even though they gave quite good results. The calculations for the tree-based models have been done after submitting the initial results and selecting another classifier, so their main function was to test the chosen classifier and compare its performance to other options that we could have chosen.

#### Random Forest \

One often used tree-based classification method is called Random Forest. It is based on the process described above, but only a random subset of the features is considered at each split of the tree. This is repeated multiple times and in the end, the obtained forest is evaluated and the best splits at each point are found. For Random Forest classifier, we used the R package randomForest[@RandomForest].

We tried the model with different parameters and with both the full dataset and the dataset with 36 variables, of which the model with all the variables performed a little better. We used that model as our Random Forest classifier in the comparison.

#### GBM \

Another tree-based model that we used was GBM (gradient boosting machine). It starts by building a set of shallow trees, but with each iteration, it learns more about the data and improves each previously built tree. By finally combining the trees, it can produce very good models for all types of data sets. It's advantage is that it doesn't necessarily require any pre-processing of the data, although the model can be improved by this. For GBM classifier, we used the R library gbm[@gbm].

We tried also this method with both the full dataset and the 168 dataset, which both gave very similar result. The model built with the 168 dataset, however, performed slightly better, so we selected that as the one to include in the comparison. Further pre-processing and variable selection didn't improve the model. We tried the model with different parameters, but in the end, couldn't produce better classifier than the ones we already had.

@@

The results for the two-class models in terms of loo-cross-validated accuracy and perplexity are:
\begin{table}
\begin{tabular}{|r|ccc|}
\hline
Model type & Loo-CV accuracy & Loo-CV perplexity&Binary accuracy on test data\\
\hline
Dummy model & 0 & 2.01&\\
Logistic regression & 0.8728448 & 1.338286&0.8445596\\
Naive Bayes & 0.8125 & 166.084 & 0.8 \\
LDA & 0.8922414 & 1.32923 &  0.8715026\\
QDA & 0.8534483 & 6.34633 & 0.8455959\\
SVM & 0.8793103 & -&0.8435233\\
Lasso & 0.8965517 & 1.29399&0.8746114\\
Random Forest & 0.887931 & 1.329754 & 0.8673575\\
GBM &	0.8943966 & 1.324385 & 0.8746114 \\
\hline
\end{tabular}
\end{table}

We also calculated for curiosity's sake the binary accuracies on the npf_test data on some of our models as can be seen above, even though these results were not part of the process of choosing our model.

@@

Our best-performing models were the the Random Forest model, the GBM model, the Lasso regression model and the linear discriminant analysis with very small differences in the binary accuracy. Out of these models, we eventually chose to use the Lasso model as it performed slightly better than the other mentioned models. The amount of information we could gain through CV calculations in R without R crashing were relatively limited, to the degree that we couldn't usually even calculate the cross-validated best choice for the tuning parameter. As a result, we chose to forego cross-validation in relation to our multinomial model.

We also chose not to use the LDA model after assessing the assumptions related to the model. LDA models have been found to perform best with a limited amount of variables and we eventually questioned, whether we should attempt to reduce the amount of variables included in the model further from the current amount of 38 from the model. On top of this, considering the assumption of equal variances, the boxplots we drew earlier in the report implied that class "nonevent" might have a larger variance in some variables than the class "event".

We made the decision of the model based on preliminary results, and though classification trees would represent an equally viable solution, we calculated those results only later on, and by that point had already chosen to focus on the other models. Initially, we calculated the results for the logistic regression model, but after noticing the error in the accuracy estimations, we chose to give up this model.

Thus, we were left with the Lasso model. However, we could question, whether for example specific variables need to be observed based on other research, and Lasso doesn't offer possibilities for this. Because cross-validation with a smaller amount of folds can give varying results different times it has been run, we eventually chose to run the cross-validation several times, check the interval of values received this way, and make decisions based on this interval.

@@

\begin{table}[!htbp]
\begin{tabular}{|r|c|}
\hline
Explanatory variable & Estimate of coefficient\\
\hline
Intercept&16.16886\\
CO2168.std&0.02065910\\
CO242.mean&-0.04024650\\
CO2504.mean&-0.02066992\\
H2O42.mean&-0.2627714\\
H2O84.std&-2.407117e-01\\
NO168.std&2.281875\\
NOx672.std&-0.06064781\\
O3672.mean&0.09233133\\
O384.std&0.03595632\\
Pamb0.std&0.2960355\\
PTG.mean&-15.19914\\
RGlob.mean&0.001578964\\
RHIRGA168.std&0.01765280\\
RHIRGA336.mean&-0.02661884\\
RHIRGA672.mean&-0.03528919\\
RHIRGA672.std&0.06932797\\
SO2168.std&0.7092859\\
SWS.mean&0.01112000\\
T168.std&0.4947142\\
CS.mean&-938.8368\\
CS.std&580.6009\\
\hline
\end{tabular}
\end{table}

@@

## Deliverables

**For the deliverables, we give a grade 4.**
